---
title: "Consumption vs Transmission (mtcars)"
author: "geotsa"
date: "October 4, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, message = FALSE)
```


## Introduction
For the needs of Motor Trend (magazine about the automobile industry) we’re looking at a data set of a collection of cars
[mtcars: (Package datasets version 3.6.1 Index)], in order to explore the relationship between a set of variables and miles
per gallon (MPG) (outcome). We are particularly interested in the following two questions:

- Is an automatic or manual transmission better for MPG?
- Quantify the MPG difference between automatic and manual transmissions?

---

## Executive Summary
If we consider the transmission type as the only regressor of mpg, then the difference between AT and MT is 7.24 mpg in
favour of the latter. Having fitted the simplest linear model, this difference is in reality the average mpg difference for the
two cases.

However, this model has major weaknesses. In addition to explaining only one third of the variation, it is characterized by
extremely high bias since, except for consumption and speed type, significant correlation exists between all the pairs of
variables parameters, and of course both mpg and am, including of course both mpg and am. The mpg regression against
am thus conceals and conceals all other significant correlations. Thus, the regression of mpg against am covers/hides all
these other significant correlations.

Among all these, most important ultimately emerge the ones between the fuel consumption and the car’s weight and
acceleration (secondarily between the fuel consumption and car’s displacement and horsepower). But including also these
regressors, the conclusions about fuel consumption and transmission mode are changing. In the case of weight alone, the
transmission mode does not seem to differentiate fuel consumption. The same happens in the case of the prediction model
that includs all other variables as regressors of mpg. Finally in the model that we include as regressor the acceleration
-in addition to am and weight- we observe that manual transmission results in (for constant weight and acceleration) a
statistically significant reduction of 2.5 mpg (increase in consumption). Hence we cannot conclude that an automatic
transmission is better for MPG than a manual one.

**Hence, we cannot conclude that an automatic transmission is better for MPG than a manual one.**

---

## Loading and preprocessing the data and the necessary libraries
```{r}
# Loading necessary libraries
library(dplyr); library(car); library(ggplot2); library(knitr); library(kableExtra)
```

```{r}
# Loading data
data("mtcars")
# Display the structure of the data (data type, variables classes)
str(mtcars)
```

The object mtcars is already a data frame of 32 observations and 11 variables. We factorize the transmission (am) variable
changing also its levels from 0/1 to AM/MT
```{r}
mtcars$am <- as.factor(mtcars$am)
levels(mtcars$am)=c("AT","MT")
```

---

## Exploratory Data Analysis
Display how many of the obs are AM and MT
```{r}
table(mtcars$am)
```
We’re using dplyr piping, group_by and summarize to calculate the mean consumption for the AM and the MT cases
```{r}
mtcars %>% group_by(am) %>% summarise(mean = mean(mpg))
```

The difference in the mean mpg values for AT and for MT seems alreay important. We ’re going a little beyond to calculate
the 6 number summaries of the mpg values for AT and for MT
```{r}
summary(mtcars[which(mtcars$am=="AT"),]$mpg)
```
```{r}
summary(mtcars[which(mtcars$am=="MT"),]$mpg)
```
The new findings also indicate a strong relationhip between the type of transmission (am) and the fuel consumtion (mpg)
[A boxplot of these summaries can be found in Appendix A].
```{r}
t.test(mtcars[mtcars$am=="AT",]$mpg, mtcars[mtcars$am=="MT",]$mpg)
```
So, for a confidence interval of 95% (a=0.05), the p-value shows that indeed the mpg means (for AT and MT) cannot be
equal.

But before we draw conclusions, we should check the correlation of all the variables to each other, and especially against
mpg. Hence, if we see the pairs of plots of Appendix B and specially its first row, we remark a clear correlation, stronger or
milder, negative or positive, between everey pair of our variables (and specially between mpg and the other variables).

Regression analysis will help us to clarify and quantify the relation between fuel consumption and transmission type.

---

## Regression Analysis

**mpg against am**

First we are fitting a model mpg~am
```{r}
fit1 <- lm(mpg~am, mtcars)
coef(fit1)
```
The model fit1 indicates that (for all the other variables fixed) the MT provokes a 7.24 mpg increase comparing to the AT.

\ 

**mpg against all the variables**

But the important correlations of the pairs of all the other variables leads us to examine the coefficients of a model with all
of them as regressors
```{r}
fit10 <- lm(mpg~., mtcars)
coef(fit10)
```
This time, the model fit10 indicates that (for all the other variables fixed) the MT provokes a 2.52 mpg increase comparing
to the AT. From the same coefficients (and from the corresponding variables’ means: Appendix C) we guess that except
transimission type it would be the displacement (disp), the rear axle ratio (drat), the horsepower (hp), the weight (wt),
number of f/w gears (gear) and the acceleration (qsec)

\ 

**Analysis of Variance**

In order to simplify our model and keep the least possible regressors, we are proceeding to a nested analysis of variance
(ANOVA). In other words we are analysing if there is a (statistically) significant difference in the mpg mean between each
model and its previous one.
```{r}
# We fit the rest 8 models
fit2 <- lm(mpg~am+disp, mtcars)
fit3 <- lm(mpg~am+disp+hp, mtcars)
fit4 <- lm(mpg~am+disp+hp+drat, mtcars)
fit5 <- lm(mpg~am+disp+hp+drat+wt, mtcars)
fit6 <- lm(mpg~am+disp+hp+drat+wt+qsec, mtcars)
fit7 <- lm(mpg~am+disp+hp+drat+wt+qsec+vs, mtcars)
fit8 <- lm(mpg~am+disp+hp+drat+wt+qsec+vs+cyl, mtcars)
fit9 <- lm(mpg~am+disp+hp+drat+wt+qsec+vs+cyl+gear, mtcars)
summary(fit9)
```

```{r}
# and we proceed to the analysis of variance of the ten total nested models
anova(fit1, fit2, fit3, fit4, fit5, fit6, fit7, fit8, fit9, fit10)
```

This analysis shows that the most significant regressors (except am that interests us) are the disp, the hp, the wt and
the qsec (in some accordance with our previous remarks). The addition of these regressors gives statistically significant
differences in the mean of the prediction, comparing to their omission.

Appendix D shows that, among them, mainly the wt (weight), but also, the qsec (accelaration) are very important to
prevent the bias that comes into the model mpg~am.

---

## Models Selection
**mpg ~ am (mdl1)**
```{r}
summary(mdl1 <- lm(mpg ~ am, mtcars))
```
According to the mdl1 model, which explains 36% of the variation, in MT cars the mpg price increases by 7.245 units
(consumption reduction). This increase is considered statistically significant (Pr<|t|). Looking at the residual plots
(Appendix E.1.), we observe that there are few ouliers available on the dataset but nothing significantly skews the data
(residuals vs. leverage). No pattern is seen across the residuals, which is good for our model.

\ 

**mpg ~ am + wt (mdl2)**
```{r}
summary(mdl2 <- lm(mpg ~ am + wt, mtcars))
```

For a constant weight, the mdl2 model —which explains this time 75% of the variation of mpg— predicts for MT cars
a slight reduction in the mpg value by 0.024 mpg (slight increase in consumption). However, this increase is considered
not statistically significant (Pr>|t|). Looking at the residual plots (Appendix E.2.), we observe that there are few ouliers
available on the dataset but nothing significantly skews the data (residuals vs. leverage). No pattern is seen across the
residuals, which is good for our model.

\ 

**mpg ~ am + wt + qsec (mdl3)**
```{r}
summary(mdl3 <- lm(mpg ~ am + wt + qsec, mtcars))
```

Finally, the mdl3 model, that explains 85% of the variation of mpg, keeping weight and acceleration constant, indicates
that MT cars have a mpg value decrease of 2.936 mpg (increase in consumption) in comparison with the AT cars. This
increase is considered statistically significant (Pr<|t|). Looking at the residual plots (Appendix E.3.), we observe that
there are few ouliers available on the dataset but nothing significantly skews the data (residuals vs. leverage). No pattern
is seen across the residuals, which is good for our model.

---
\ 

\ 

\ 

\ 

\ 

# Appendix

\ 

#### A. MPG: AT vs MT
```{r}
ggplot(mtcars, aes(am,mpg)) + geom_boxplot() + labs(x= "Transmission (Automatic/Manual)",y = "Fuel Consumption (mpg=Miles/Gallon)")
```

#### B. CORRELATIONS
```{r}
corr <- select(mtcars, mpg,cyl,disp,hp, drat,wt,qsec,vs,am,gear,carb)
pairs(corr)
```

#### C. COEFFICIENTS x MEAN
```{r}
Mean <- sapply(select(mtcars,-c(mpg,am)),mean)
Coeff <- c(coef(fit10)[2:8],coef(fit10)[10:11])
.x. <- Mean*Coeff
round(rbind(Mean, Coeff, .x.),3)
```

#### D. ANOVA - Comparison of all the possible nested models - Average Pr(>F)
We are creating nested models so as each one of the four regressors to be added/included in all the four possible steps and
we take the average of their Pr.
```{r}
anova(fit1, fit2, fit3, fit5, fit6)
```

```{r}
fi2 <- lm(mpg~am+hp, mtcars)
fi3 <- lm(mpg~am+hp+wt, mtcars)
fi5 <- lm(mpg~am+hp+wt+qsec, mtcars)
fi6 <- lm(mpg~am+hp+wt+qsec+disp, mtcars)
anova(fit1, fi2, fi3, fi5, fi6)
```

```{r}
fii2 <- lm(mpg~am+wt, mtcars)
fii3 <- lm(mpg~am+wt+qsec, mtcars)
fii5 <- lm(mpg~am+wt+qsec+disp, mtcars)
fii6 <- lm(mpg~am+wt+qsec+disp+hp, mtcars)
anova(fit1, fii2, fii3, fii5, fii6)
```

```{r}
fiii2 <- lm(mpg~am+qsec, mtcars)
fiii3 <- lm(mpg~am+qsec+disp, mtcars)
fiii5 <- lm(mpg~am+qsec+disp+hp, mtcars)
fiii6 <- lm(mpg~am+qsec+disp+hp+wt, mtcars)
anova(fit1, fiii2, fiii3, fiii5, fiii6)
```

```{r}
disp <- round((anova(fit1, fit2, fit3, fit5, fit6)$Pr[2]+
anova(fit1, fi2, fi3, fi5, fi6)$Pr[5]+
anova(fit1, fii2, fii3, fii5, fii6)$Pr[4]+
anova(fit1, fiii2, fiii3, fiii5, fiii6)$Pr[3])/4,4)
hp <- round((anova(fit1, fit2, fit3, fit5, fit6)$Pr[3]+
anova(fit1, fi2, fi3, fi5, fi6)$Pr[2]+
anova(fit1, fii2, fii3, fii5, fii6)$Pr[5]+
anova(fit1, fiii2, fiii3, fiii5, fiii6)$Pr[4])/4,4)
wt <- round((anova(fit1, fit2, fit3, fit5, fit6)$Pr[4]+
anova(fit1, fi2, fi3, fi5, fi6)$Pr[3]+
anova(fit1, fii2, fii3, fii5, fii6)$Pr[2]+
anova(fit1, fiii2, fiii3, fiii5, fiii6)$Pr[5])/4,4)
qsec <- round((anova(fit1, fit2, fit3, fit5, fit6)$Pr[5]+
anova(fit1, fi2, fi3, fi5, fi6)$Pr[4]+
anova(fit1, fii2, fii3, fii5, fii6)$Pr[3]+
anova(fit1, fiii2, fiii3, fiii5, fiii6)$Pr[2])/4,4)
#kable_styling(kable(rbind(c("","disp","hp","wt","qsec"),c("mean(Pr(>F))",disp,hp,wt,qsec)),
#                    format = "latex"))
```

#### E. MODEL RESIDUALS

**1. mpg ~ am**
```{r}
par(mfrow = c(2,2)); plot(mdl1)
```

**2. mpg ~ am + wt**
```{r}
par(mfrow = c(2,2)); plot(mdl2)
```

**3. mpg ~ am + wt + qsec**
```{r}
par(mfrow = c(2,2)); plot(mdl3)
```